A radiocarbon data based evaluation of Cultural Transmission in Bronze Age burial rites of Central, Northern and North-western Europe

### 2019-01-20 article/04_Discussion.Rmd
some possible interpretation of the results

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation can be assumed based on the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000calBC) and a decline (after 2000) of cultural-spatial-distance correlation. One interpretation of this oberservation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. There are two main reasons why such interpretations are difficult so far:

1. The amount, distribution and quality of the data as discussed in section \ref{data-quality-discussion} queries the credibility of the cultural distance calculation.

2. Data meaning \ref{data-meaning-discussion}

Only two variables

Keeping this in mind, the surprising result is still valid and has to be considered in future hypotheses building.

>>>>

Against this background, the observation that there is no permanent, significant correlation either between the variables or with the spatial distance is all the more astonishing (Table \ref{tab:distanceandcorrelationtable*}). Only for small periods of time this correlation can be assumed based on the Radon-B data. The *burial construction* distance shows a higher degree of correlation with spatial distance than the *burial type* distance, which even seems to behave in an opposing manner: negative correlation. If one assumes the reconstructed timelines of cultural similarity intensity to be correct, than one has to explain a rise (between 2200 and 2000calBC) and a decline (after 2000) of cultural-spatial-distance correlation. There are several reasons why any interpretation of this observation is difficult so far:

1. The amount, distribution and quality of the data as discussed in section \ref{data-quality-discussion} queries the credibility of the cultural distance calculation. The macroregions have to be very big due to a lack of data and don't allow for a precise picture.

2. The data used for this analysis was collected for another purpose and its application in this context requires a problematic shift in data meaning as explained in section \ref{data-meaning-discussion}.

3. All observations of cultural distance in this paper are based on only the two variables *burial type* and *burial construction*. This is by far not enough for any overarching conclusions about the general state of society in the Bronze Age.

Keeping this in mind, the surprising result is still valid and has to be explained. One interpretation might be, that the Early Bronze Age was a time of rising cultural exchange and religious homogenization, while in later periods different domains secluded themselves more strongly. The negative correlation of *burial type* in the Late Bronze Age could indicate long distance contacts that explicitly excluded certain intermediate regions of Central Europe. The rise of the Tumulus Culture in the Middle Bronze Age might have cushioned this effect for the variable *burial construction* and at the same time introduced more variability for a less important aspect *burial type*. The latter may have been of subordinate relevance in the Tumulus Culture context. With the rise of the Urnfield Culture this phenomonon could have been reversed as *burial type* became an expression of something that could have been one of the first formalized religions in Europe: The spread of the Urnfield Culture was an extraordinary development in early European history and was possibly accompanied by a structured, religious dogma that could have built up around the symbolic content of cult wagons, sunbarks and waterfowl [@bilicSwanChariotSolar2016; @kimmig_seevolkerbewegung_1964; @molloy_european_2018]. Of course these explanations could, if accurate, only be considered on a really big scale. For individual regions and sites specific ideological, social or even economic considerations might have been much more critical for the adopted burial tradition [@barrett_monumentality_1990; @fokkens_genesis_1997-1; @muller-scheesel_variabilitat_2009].
